---
title: "Assignment 3: Data Exploration"
author: "Samantha Burch"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 28 at 1:00 pm.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 
```{r,  message = FALSE}
getwd()
library(tidyverse)

read.csv(file.choose(ECOTOX_Neonicotinoids_Insects_raw))
Neonics <- read.csv(file.choose(ECOTOX_Neonicotinoids_Insects_raw))

Litter <- read.csv(file.choose(NEON_NIWO_Litter_massdata_2018_08_raw))
View(Litter)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotoids are considered to be highly effective insecticides for the following: 1) crop protection against pests, and 2) flea control for both cats and dogs. These widly spread/used insecticides could be posing a threat to aquatic environments, as they first contaminate the soil on which they're used and then their residues trickle down into our water systems. To date, there is little awareness of the impacts of Neonicotoids on aquatic environments and ecosystems overall; thus, it is important to close such mentioned knowledge gaps by further analyzing data linked to its use.(https://www.ncbi.nlm.nih.gov/pubmed/15822177)

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: It is important to study this as such litter and woody debris can impact forest communities and negatively impact ground-dwelling invertebrates. These types of disturbances can alter habitat structure, energy and nutrient flow, and ultimately shape critical ecosystem processes. (https://www.mdpi.com/1999-4907/8/5/174/htm)

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Each collection event is measured separately for different functional groups (i.e. leaves, twigs, needles) 
* All masses are reported "at the spatial resolution of a single trap and the temporal resolution of a single collection event." No single site should have more than 3,440 data instances in a single calendar year.
* It is important to pay attention to data relationships and ensure to check the data for anomolies before joining tables. Lastly, in order to provide context to litter data, users are encouraged to leverage data from vegetation structure.


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(ECOTOX_Neonicotinoids_Insects_raw)
```

6. Using the `summary` function, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(ECOTOX_Neonicotinoids_Insects_raw)
```

> Answer: It is critical to understand if there are any discrepancies in, say, how long an organism was observed and its specific exposure type. This also allows you to see important summary statistics in one place, including for example: mean, median, and quartiles.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}

summary(Neonics)

```

> Answer: In this dataset, the six most commonly studied species include: 1) Honeybee, 2) Parasitic Wasp, 3) Carniolan Honeybee, 4) Bumble Bee, and 6) Italian Honeybee. These mentioned species can be categorized into 'pollinator insects,' which indicates that they are key to the vitality of our ecosystems.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class("Con.1.Author")
```

> Answer: This is considered to be a character class. It is not numeric because it is listed as an active ingredient.


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics)

ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, binwidth = 15, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The two most common publications and test locations are via 1) the lab, and 2) field natural. These generally peak and dip at the same time; however, there is a large jump in lab use between about 2013 and 2015. This could denote years in which there were an increased number of publications and lab space became more readily available for conducting tests.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r, fig.height = 3, fig.width = 4}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()

class("Endpoint")
```

> Answer: The two most common endpoints are 'NOEL' and 'LOEL,' and they are defined as characters.
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class("collectDate") #character
class(Litter$datetime)

Litter$datetime <- format(Litter$datetime, "%y")
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)

unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID) #12
```

> Answer: The summary function is considered a generic function that produces summaries of the results of "various model fitting functions." Where as the unique function is used to return a vector, data frame or array with any duplicate rows/elements removed.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: __ had the highest biomass at these sites.