---
title: "Assignment 5: Data Visualization"
author: "Samantha Burch"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 11 at 1:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse and cowplot packages. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (tidy and gathered) and the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
#load tidyverse and cowplot packages; upload NTL-LTER processed data files (Peter and Paul) and processed data file for Niwot
library(tidyverse)
library(cowplot)
library(ggplot2)

TIDYPeterPaulNut <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")
TIDYPeterPaulChemPhys <- read.csv("./Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
NiwotLitter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
class(TIDYPeterPaulNut$sampledate)

TIDYPeterPaulNut$sampledate <- as.Date(TIDYPeterPaulNut$sampledate, format = "%Y-%m-%d")
View(TIDYPeterPaulNut$sampledate)


```
## Define your theme
3. Build a theme and set it as your default theme.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "blue"), 
        legend.position = "top")

theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus by phosphate, with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
library(viridisLite)
library(viridis)
library(RColorBrewer)
library(colormap)


NvsP <-
  ggplot(TIDYPeterPaulNut, aes(x = tp_ug, y = po4)) +
  geom_point(aes(color = lakename)) + #have to have aesthetics bc it was looking for a color named lakename, which wouldn't work but works with aesthetic
  geom_smooth(method = lm, color = "black") +
  labs(x = "Phosphorous", y = "Phosphate",
       color = "Depth (m)", shape = "")
print(NvsP)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}

#5-1 Temperature Boxplot
Tempboxplot <-
  ggplot(TIDYPeterPaulNut, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(x = expression(paste("Month")),
       y = expression(paste("Temperature" (degree~C))), color = "Lake") + 
  scale_y_continuous(expand = c(0,0)) +
  scale_color_viridis(discrete = TRUE, option = "viridis", end = 0.8) +
  theme(legend.position = "none")
print(Tempboxplot)

#5-2 TP Boxplot
TPboxplot <-
  ggplot(TIDYPeterPaulNut, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(x = expression(paste("Month")),
       y = expression(paste("TP" ~ (mu*g / L))), color = "Lake") + 
  scale_y_continuous(expand = c(0,0)) +
  scale_color_viridis(discrete = TRUE, option = "viridis", end = 0.8) +
  theme(legend.position = "none")
print(TPboxplot)

#5-3 TN Boxplot
TNboxplot <-
  ggplot(TIDYPeterPaulNut, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(x = expression(paste("Month")),
       y = expression(paste("TN" ~ (mu*g / L))), color = "Lake") + 
  scale_y_continuous(expand = c(0,0)) +
  scale_color_viridis(discrete = TRUE, option = "viridis", end = 0.8) +
  theme(legend.position = "none")
print(TNboxplot)

#5-4 Cowplot that combines all three

threeplots <- plot_grid(Tempboxplot, TPboxplot, TNboxplot, Labels = c('Temp', 'TP', 'TN'),
                        label_size = 10, align = "vh", ncol = 1)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}

#6
NIWOneedles <- 
  ggplot(subset(NiwotLitter, functionalGroup == "Needles"), #why double equal signs?
         aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point()

print(NIWOneedles) #how to fix x axis so I can see all collectDates?

#7
NIWOneedlesFacets <- 
  ggplot(subset(NiwotLitter, functionalGroup == "Needles"), #why double equal signs?
         aes(x = collectDate, y = dryMass) +
  geom_point() +
    facet_wrap(vars(nlcdClass), nrow = 3)
  
print(NIWOneedlesFacets)

```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: __ is more effective because