---
title: "Assignment 2: Coding Basics"
author: "Samantha Burch"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. seq(1, 100, 4) # 1-100 by 4 ans: [1]  1  5  9 13 17 21 25 29 33 37 41 45 49 53 57 61 65 69 73 77 81 85 89 93 97
seq1 <- seq(1, 100, 4)
seq1

#2. 
mean(seq(1, 100, 4)) # ans: 49
median(seq(1, 100, 4)) # ans: 49

#3. 
mean(seq(1, 100, 4))>median(seq(1, 100, 4)) # ans: FALSE
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5 and 6
vectorstudents <- c("Sam", "Carlen", "Rachel", "Ian") # character vector
vectorstudents

vectorgrades <- c(100,98,92,85) # numeric vector 
vectorgrades

vectorpassfail <- c(TRUE, TRUE, TRUE, TRUE) # logical vector
vectorpassfail

#7
dataframegrades <- data.frame(vectorstudents, vectorgrades, vectorpassfail)
dataframegrades #created data frame using each of my named vectors

#8
names(dataframegrades) <- c("Students", "Grades", "Pass/Fail"); View (dataframegrades)
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is different from a matrix in that each column can be assigned a different mode (i.e. numbers in one, characters in another). Matrices only show one element/the same type of data.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10 (if and else)
x <- 50
x
grade <- function(x) {
  if(x < 50) {x = FALSE}
  else {
    x = TRUE
  }
}

#ifelse
y <- 50
y
grade2 <- function(y) {
  ifelse(y<50, FALSE, TRUE)
}
grade2 #TRUE, TRUE, TRUE, TRUE

#11: results
Sam <- grade(100); Sam
Samifelse <- grade(100); Samifelse

Carlen <- grade(98); Carlen
Carlenifelse <- grade(98); Carlenifelse

Rachel <- grade(92); Rachel
Rachelifelse <- grade(92); Rachelifelse

Ian <- grade(85); Ian
Ianifelse <- grade(85); Ianifelse

grade <- grade(vectorgrades)
grade2 <- grade2(vectorgrades)
grade2
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:  The 'if' and 'else' option worked, as I was able to apply my vector function to each individual student to determine a passing grade - these outputs appeared in both the console and the 'Environment' under values. When I ran the ifelse option, the following message appeared: "Warning message: In if (x < 50) { : the condition has length > 1 and only the first element will be used." However, when I run 'grade2' on its own, I do see the following output: "[1] TRUE TRUE TRUE TRUE." That said, it allows me to see all answers at once.
